Example script the `dispRity` paper

### 2018-03-16 paper/dispRity_paper.tex

\usepackage{xcolor}
\usepackage{color,soul}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\usepackage{hyphenat}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\newcolumntype{d}[1]{D{.}{.}{#1}}

@@

\noindent {\normalsize \sc Thomas Guillerme$^12$$^,$$^*$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}

@@

        \item There are as many definitions of these multidimensional spaces, metrics and tests as there are questions that can be tackled with such methods.
        Many of these methods are implemented in specific software or R packages.
        However, the definition of the space, metric and test is often dependent on the software/package's author's point of view or specific question.
        This can unfortunately hamper researchers' ability to apply different methods that best suits their specific questions.

        \item Here I present the \disp package, a flexible R package for running multidimensional analysis.
        This package allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
        It also provides a tidy interface through the \disp object, allowing users to easily run reproducible multivariate analysis.

        \item The \disp package also comes with an extend manual regularly updated following users' questions or suggestions.
        Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
        Finally, it also contains a suite of utilities to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.

@@

% Multidimensionality
Biological data are complex.
To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g. \citealt{Close2015}; and continuous traits such as lengths, e.g. \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.

In all these analysis, each set of multivariate traits forms a multidimensional space.
This space is represented as a matrix where rows are regarded as samples or observations (e.g. a specimen, field sites, etc.) and columns are variables or some transformation thereof (e.g. an embedding, scaling, ordination, etc. of the variables).
These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (\citealt{lloyd2016estimating} and references therein; e.g. in \citealt{Close2015}), or as outputs from an ordination, whether it being a principal components analysis (PCA,
\citealt{PCA}; e.g. in \citealt{zelditch2012geometric} ), a metric scaling (PCO, PCoA, \citealt{PCO} ; e.g. in \citealt{Brusatte12092008} ) or a non-metric scaling (MDS, NMDS, \citealt{MDS}
; e.g. in \citealt{Liow2004,DonohueDim}).
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits, the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.

@@

Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (
NPMANOVA, \citealt{NPMANOVA}; e.g. in \citealt{Brusatte12092008}) multidimensional permutation tests ( \citealt{ManlyPermutations}; e.g. in \citealt{diaz2016global}) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
This is ideal for testing volume based hypothesis (e.g. does groups A and B have the same volume?), but in other cases may not be appropriate in non-volume based hypothesis, e.g. does groups A and B occupy the same location?).
This can lead to inappropriate analyses by users confined by the existing software implementations.

The aim of the \disp package is avoid such problems by providing a completely flexible framework for studying multidimensional data.
This package is based on a highly modular architecture where each crucial decision in multidimensional analysis (which data, which metric and which test) can be provided by the user.
It implements many commonly used metrics and definitions of disparity, as well as providing a simple interface for users to implement their own disparity functions.
The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data
(see the glossary Table \ref{Tab:terms} for terms' equivalences between palaeobiology and ecology).

\begin{landscape}
\begin{table}
    \begin{tabular}{p{4cm}p{5cm}p{6cm}p{6cm}}
        \hline
        \textbf{in this manuscript} & \textbf{in} \texttt{\textbf{dispRity}} & \textbf{in palaeobiology} & \textbf{in ecology} \\ 
        \hline
        multidimensional space & matrix ($n\times k$) & morphospace, traitspace, etc. & ecospace, function-space, etc. \\
        \hline
        elements & rows ($n$) & taxa, specimen, etc. & taxa, field sites, environments, etc. \\
        \hline
        dimensions & columns ($k$) & ordination scores, distances, etc. & ordination scores, distances, etc. \\
        \hline
        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in a stratum or sharing the same ancestor & e.g. elements living in the same environment \\
        \hline
        disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g. the sum of variances \cite{Wills2001}, the average pairwise distances between taxa \cite{Foote01071994}, etc. & dissimilarity: e.g. ellipsoid volume \cite{DonohueDim}, convex hull volume \cite{cornwell2006trait}, etc. \\
        \hline
    \end{tabular}
    \caption{Glossary and equivalences between this manuscript, the \texttt{dispRity} package and terms commonly used in palaeobiology or ecology. $^1$See Fig. \ref{Fig:levels}.}
    \label{Tab:terms}
\end{table}
\end{landscape}

\section{Description}
In brief, the package takes a \texttt{matrix} object (the multidimensional space), 
calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation.
Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements \citep[e.g. the space can be a distance matrix][]{Close2015}.
The matrix is always considered as the multidimensional space and no correction is applied to it (e.g. if ordinated and with negative eigen values, corrections should be applied prior to using the \texttt{dispRity} package.)

@@

\subsection{Measuring disparity}
The \disp function measures disparity of any multidimensional space (of class \texttt{matrix} or \texttt{data.frame}) where the columns correspond to the dimensions and the rows correspond to the elements present in the space.
The disparity metric (i.e. the aspect of interest of the space) is passed through the \texttt{metric} argument and is defined by the user as any function or combination of functions that can either transform the matrix into:

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{../inst/gitbook/dispRity_fun.pdf} 
\caption{Illustration of the different metric dimension-levels in the \disp package. In this example, each cell corresponds to a single value (e.g. a 8 $\times$ 7 matrix or a vector of eight elements).
A dimension-level 3 matrix would be a metric resulting in a matrix (e.g. the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would result in a vector (i.e. a distribution, e.g. \texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would result in a single value (e.g. \texttt{stats::sd} which calculates the standard deviation of the input matrix)}.
\label{Fig:levels}
\end{figure}

The disparity metrics can be functions from other packages, defined by the user or from the \disp package (see Table \ref{Tab:metrics}).
When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
For example, if the metric is defined as \texttt{metric = c(prod, ranges)} (the hypercube volume), the \texttt{ranges} function (dimension-level 2) is first applied to data (the range of each dimensions) and the function \texttt{prod} is then applied to the results (\texttt{prod(ranges(data))}).
One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).
Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).

\begin{table}
\resizebox{\textwidth}{!}
{%
    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
        \textbf{name} & \textbf{description} & \textbf{dim} & \textbf{definition} & \textbf{source} \\
        \hline
        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
        \hline
        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
        \hline
        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \hline
        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \hline
        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
        \hline
        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
        \hline
        \texttt{mode.val} & The modal value & 1 & NA & This package\\
        \hline
        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
        \hline
        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
        \hline
        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
        \hline
        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
        \hline
        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
        \hline
        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
    \end{tabular}
}%
    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
    \label{Tab:metrics}
\end{table}

\textbf{Table 1 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
\label{Tab:metrics}

\subsection{Splitting the multidimensional space into subsets}
Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
For example, one may be interested in seeing how the disparity of a specific subset of the space compares to another (e.g. difference among groups) or, how different subsets change sequentially (e.g. through time).
In essence, the original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
Conversely, subsets correspond to parts of the space with pooled characteristics.

This splitting can be done using the \texttt{custom.subsets} or \texttt{chrono.subsets} functions.
The first argument is the matrix defining the space and a list of elements to group into different subsets.
The second also takes a matrix and arguments giving the age of the taxa (a dated phylogeny of the elements present in the morphospace - see below) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice.
For each method, the selected data is:

\begin{itemize}
    \item ``acctran'': always the descendant
    \item ``deltran'': always the ancestor
    \item ``random'': randomly chosen between the descendant or the ancestor
    \item ``proximity'': either the descendant or the ancestor depending on branch length
    \item ``equal.split'': both the descendant and the ancestor with an even probability
    \item ``gradual.split'': both the descendant and the ancestor with a probability depending on branch length
\end{itemize}

Note that the four first models are a proxy for punctuated evolution and the two last ones for gradual evolution.
There is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.
See Guillerme \& Cooper \textit{in review} for full description.

@@

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms: (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data); and (2) the ``single'' algorithm where $n = 1$ (somewhat akin to jackknife resampling).

@@

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
Additionally, the \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

>>>>

\usepackage{xcolor}
\usepackage{color,soul}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\usepackage{hyphenat}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{float}
\restylefloat{table}
\newcolumntype{d}[1]{D{.}{.}{#1}}

@@

\noindent {\normalsize \sc Thomas Guillerme$^{1,2,*}$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\
$^2$School of Biological Sciences, University of Queensland, St. Lucia, Australia.\\}

@@

        \item There are as many definitions of these multidimensional spaces, metrics and tests as there are questions that can be tackled with such methods.
        Many of these methods are implemented in specific software or R packages.
        However, the definition of the space, metric and test is often dependent on the software/package authors points of view or specific questions.
        This can unfortunately hamper researchers' ability to apply different methods that best suits their specific questions.

        \item Here I present the \disp package, a flexible \texttt{R} package for performing multidimensional analysis.
        It allows users to define each step of the analysis (whether it is the space, the metric or the test) through a highly modular architecture where each definition can be passed as a function.
        It also provides a tidy interface through the \disp object, allowing users to easily run reproducible multivariate analysis.

        \item The \disp package also comes with an extend manual regularly updated following users' questions or suggestions.
        Furthermore, the package contains some simulation tools (e.g to simulate complex multidimensional space or morphological data).
        Finally, it also contains a suite of utility functions to work with \disp objects aimed at helping users to develop their own multidimensional metrics and/or tests.

@@

% Multidimensionality
Biological data are complex.
To understand the ecology and evolution of species we must use multiple variables that inevitably covary with each other through time and space.
One solution to this problem is to analyse these data in a multivariate framework \citep[e.g.][]{price2015predation,diaz2016global}.
Such analyses aim to capture the complex multidimensionality of biological data, while still providing outputs that are interpretable.
These multivariate analyses can be used to investigate changes in morphological diversity through time \citep[e.g.][]{Close2015}, competitive replacement scenarios \citep[e.g.][]{Brusatte12092008}, relationships among form and function \citep[e.g.][]{diaz2016global} and even to describe the entirety of possible shapes for a group of organisms \citep[e.g.][]{raup1966geometric}.
The biological variables in such analyses are equally diverse, including morphological traits (discrete traits like the presence or absence of a character, e.g. \citealt{Close2015}; or continuous traits such as lengths, e.g. \citealt{price2015predation}), life history traits \citep[e.g.][]{diaz2016global}, or even ecosystem properties \citep[e.g.][]{DonohueDim}.

In all these analysis, each set of multivariate traits forms a multidimensional space.
This space is represented as a matrix where rows are regarded as samples or observations (e.g. specimens, field sites, etc.) and columns are variables or some transformation thereof (e.g. embedding, scaling, ordination, etc.).
These multidimensional spaces can be defined in many ways, for example as a pairwise distance matrix (\citealt{lloyd2016estimating} and references therein; e.g. in \citealt{Close2015}), or as outputs from an ordination, whether it being a principal components analysis (PCA,
\citealt{PCA}; e.g. in \citealt{zelditch2012geometric} ), a metric scaling (PCO, PCoA, \citealt{PCO} ; e.g. in \citealt{Brusatte12092008} ) or a non-metric scaling (MDS, NMDS, \citealt{MDS}
; e.g. in \citealt{Liow2004,DonohueDim}).
The name we give to the multidimensional space tends to vary with the kinds of traits used to construct it. 
For example, when using morphological traits, the space will be a morphospace, when using ecological traits it may be referred to as an ecospace, etc.

@@

Finally, all these different multidimensional spaces and their associated disparity metrics can be used in an equal variety of statistical tests such as non-parametric multivariate analyses of variance (NPMANOVA, \citealt{NPMANOVA}; e.g. in \citealt{Brusatte12092008}) multidimensional permutation tests (\citealt{ManlyPermutations}; e.g. in \citealt{diaz2016global}) or even, less rigorously, by looking at the confidence interval overlaps between disparity measurements.
In summary, there are many different ways to perform each step of a multidimensional analysis, making analyses of complexity ever more complex.

In theory, this multitude of ways to generate and define multidimensional spaces, measure disparity within and analyse these metrics is not an issue, in fact, it allows researchers to choose both the most appropriate method for their question or data, or even to test their question using multiple methods.
In practice, however, this is hampered by existing software implementations.
Although many software packages exist for multidimensional analysis \citep[e.g.][]{navarro2003MDA,bouxin2005ginkgo,de2007ginkgo,oksanen2007vegan,geiger2008,adams2013geomorph,Claddis,adams2017geometric},
package maintainers/software developers choose their preferred definition of multidimensional space and disparity metric to best fit their needs (i.e. data, hypothesis, etc.) making the implementations sometimes hard to adapt to different needs.
For example, in the excellent and widely used \texttt{geomorph} package, morphological disparity analysis uses the \texttt{morphol.disparity} function that defines the multidimensional space as the ordination of the Procrustes transform of the morphometric data, the disparity metric
as the relative sum of the diagonal of the covariance of the ordination scores (Procrustes variance)
, and uses permutation tests \citep{zelditch2012geometric,adams2013geomorph,adams2017geometric}.
This is ideal for testing volume based hypothesis (e.g. ``does groups A and B have the same volume?''), but in other cases may not be appropriate in non-volume based hypothesis (e.g. ``do they occupy the same location?'').
This can lead to inappropriate analyses by users confined by the existing software implementations.

The aim of the \disp package is avoid such problems by providing a flexible framework for studying multidimensional data.
This package is based on a modular architecture where each decision in multidimensional analysis (which data, metric and test) can be specified by the user.
It implements many commonly used disparity metrics, as well as providing a simple interface for users to implement their own disparity functions.
The package is described here for the use of discrete morphological data disparity analysis but can be generalised to any type of multidimensional data (see the glossary Table \ref{Tab:terms}).

\begin{landscape}
\begin{table}
    \begin{tabular}{p{4cm}p{5cm}p{6cm}p{6cm}}
        \hline
        \textbf{in this manuscript} & \textbf{in} \texttt{\textbf{dispRity}} & \textbf{in palaeobiology} & \textbf{in ecology} \\ 
        \hline
        multidimensional space & matrix ($n\times k$) & morphospace, traitspace, etc. & ecospace, function-space, etc. \\
        \hline
        elements & rows ($n$) & taxa, specimen, etc. & taxa, field sites, environments, etc. \\
        \hline
        dimensions & columns ($k$) & ordination scores, distances, etc. & ordination scores, distances, etc. \\
        \hline
        subsets & matrix ($m \times k$, with $m \leq n$) & e.g. every element in a stratum or sharing the same ancestor & e.g. elements living in the same environment \\
        \hline
        disparity & a dimension-level 1 or 2 function$^1$ & disparity: e.g. the sum of variances \citep{Wills2001}, the average pairwise distances between taxa \citep{Foote01071994}, etc. & dissimilarity: e.g. ellipsoid volume \citep{DonohueDim}, convex hull volume \citep{cornwell2006trait}, etc. \\
        \hline
    \end{tabular}
    \caption{Glossary and equivalences between this manuscript, the \texttt{dispRity} package and terms commonly used in palaeobiology or ecology. $^1$See Fig. \ref{Fig:levels}.}
    \label{Tab:terms}
\end{table}
\end{landscape}

\section{Description}
In brief, the package takes a \texttt{matrix} object (the multidimensional space), calculates a disparity metric from the space and analyse the resulting \texttt{dispRity} object through hypothesis testing and visualisation.
Some additional functions modify the space, for example by dividing it by groups or through time and/or bootstrapping it (see Fig. \ref{Fig:workflow}).
Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements \citep[e.g. the space can be a distance matrix][]{Close2015}.
The matrix is always considered as the final multidimensional space to analyse and no correction is applied to it (e.g. potential corrections should be applied prior to using the \texttt{dispRity} package).

@@

\subsection{Measuring disparity}
The \disp function measures disparity from a \texttt{matrix} where the columns correspond to the dimensions and the rows correspond to the elements present in the space.
The disparity metric is passed through the \texttt{metric} argument and is defined by the user as one or more function(s) that can either transform the matrix into:

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.8\textwidth]{dispRity_fun.pdf} 
\caption{Illustration of the different metric dimension-levels in the \disp package. In this example, each cell corresponds to a single value (e.g. a 8 $\times$ 7 matrix or a vector of eight elements).
A dimension-level 3 matrix would be a metric output a matrix (e.g. the function \texttt{stats::cor} to calculate the correlation between each dimension), a dimension-level 2 metric would output a vector (i.e. a distribution, e.g. \texttt{dispRity::variances} which calculates the variance within each dimension) and a dimension-level 1 metric would output a single value (e.g. \texttt{stats::sd} which calculates the standard deviation of the input matrix)}.
\label{Fig:levels}
\end{figure}

The disparity metrics can be any \texttt{R} function (see Table 2 for metrics implemented in the package).
When multiple functions are passed to the \texttt{metric} argument, they are sorted by dimension-level and applied in decreasing order to the data.
For example, if the metric is defined as \texttt{metric = c(prod, ranges)} (the hypercube volume), the \texttt{ranges} function (dimension-level 2) is first applied to data and the function \texttt{prod} is then applied to the results (\texttt{prod(ranges(data))}).
One can also directly pass a function description to the \texttt{metric} argument (e.g. \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} for the average squared pairwise distance).
Note that this function also allows to work on only a subset of dimensions via the \texttt{dimensions} argument (e.g. if only the $n$ first dimensions must be considered).

\begin{table}[H]
\resizebox{\textwidth}{!}
{%
    \begin{tabular}{p{2cm}|p{5cm}|p{1cm}|p{5cm}|p{5cm}}
        \textbf{name} & \textbf{description} & \textbf{dim} & \textbf{definition} & \textbf{source} \\
        \hline
        \texttt{ancestral. dist} & The distance between an element and its ancestor & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Ancestor_{n})^2}}$ & This package \\
        \hline
        \texttt{centroids} & The distance between each element and a fixed point$^1$ of the space & 2 & $\sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}$ & This package \\
        \hline
        \texttt{convhull. surface} & The surface of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \hline
        \texttt{convhull. volume} & The volume of the convex hull & 1 & NA & \texttt{geometry::convhulln} \citep{barber1996quickhull,geometryR}\\
        \hline
        \texttt{diagonal} & The greatest Euclidean distance & 1 & $\sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}$ & This package \\
        \hline
        \texttt{ellipse. volume}$^2$ & The volume of the ellipsoid & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})$ & This package; based on \cite{DonohueDim}\\
        \hline
        \texttt{mode.val} & The modal value & 1 & NA & This package\\
        \hline
        \texttt{n.ball. volume} & The hyper-spherical ($n$-ball) volume & 1 & $\frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} R$ & This package\\
        \hline
        \texttt{pairwise. dist} & The pairwise distances between elements & 2 & NA & \texttt{vegan::vegdist} \citep{oksanen2007vegan}\\        
        \hline
        \texttt{radius} & The radius of each dimensions & 2 & $|\frac{\sum_{i=1}^{n}k_i}{n} - f(\mathbf{v}k)|$ & This package\\
        \hline
        \texttt{ranges} & The absolute ranges of each dimension & 2 & $|max(k_i) - min(k_i)|$ & This package \\
        \hline
        \texttt{span.tree. length} & The minimal spanning tree length & 1 & $\sum(\mathrm{branch\ length})$ & \texttt{vegan::spantree} \citep{oksanen2007vegan}\\
        \hline
        \texttt{variances} & The variance of each dimension & 2 & $\sigma^{2}{k_i}$ & This package \\
    \end{tabular}
}%
    \caption{\small{Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is a function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO/PCoA (\textit{not} PCA).}}
    \label{Tab:metrics}
\end{table}

\textbf{Table 2 (caption):} Where $k$ is the number of dimensions, $n$ the number of elements, $\Gamma$ is the Gamma distribution, $\lambda_i$ is the eigenvalue of each dimensions, $\sigma^{2}$ is their variance and $Centroid_{k}$ is their mean, $Ancestor_{n}$ is the coordinates of the ancestor of element $n$, $f(\mathbf{v}k)$ is function to select one value from the vector $\mathbf{v}$ of the dimension $k$ (e.g. it's maximum, minimum, mean, etc.), $R$ is the radius of the sphere or the product of the radii of each dimensions ($\displaystyle\prod_{i=1}^{k}R_{i}$ - for a hyper-ellipsoid). $^1$ by default that point is the centroid of the elements; $^2$ this function uses a fast estimation of the eigenvalue that only works in an ordinated space based on MDS or PCO (\textit{not} PCA).
\label{Tab:metrics}

\subsection{Splitting the multidimensional space into subsets}
Prior to calculating disparity, the space can be subdivided into subsets, typically to be compared to each other.
For example, one may compare the disparity of a specific subset of the space to another or, how different subsets change sequentially (e.g. through time).
The original space corresponds to the overall space (e.g. a morphospace contains all the observed morphologies).
Subsets correspond to parts of the space with pooled characteristics.

This splitting can be done using the \texttt{custom.subsets} or \texttt{chrono.subsets} functions.
The first function takes a matrix defining the space and a list of elements defining the subsets.
The second also takes a matrix and arguments giving the age of the taxa (a dated phylogeny of the elements present in the morphospace - see below) and which subsets to create: (1) discrete time subsets (or time-binning) or (2) continuous time subsets (or time-slicing).

The time-binning method groups elements by specific age range.
The time-slicing method works by using a phylogeny and looking at which taxa are present at any specific point in time.
This method thus requires the nodes to be part of the space, a dated phylogeny (chronogram) and which model to use when slicing through branches rather than tips and nodes.
When a slice occurs not on a tip or a node, six methods are available to select either the descendent or the ancestor's node/tip as an element for this time slice:
``acctran'', ``deltran'', ``random'' and ``proximity'' as proxy for punctuated evolution models; and ``equal.split'', ``gradual.split'' as a proxy for gradual evolution.
See \cite{time-slice} for full description of the method.
Note that there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also decreases accuracy.

@@

Bootstrapping can be achieved by using the \texttt{boot.matrix} function which pseudo-replicates the space following two algorithms: (1) the ``full'' algorithm where the bootstrapping is entirely stochastic ($n$ elements are replaced by any $m$ elements drawn from the data); and (2) the ``single'' algorithm where $n = 1$ (similar to jackknife).

@@

\subsubsection{Summarising and plotting}
The \texttt{summary.dispRity} and \texttt{plot.dispRity} functions allow users to set which central tendency and which quantiles should be represented.
The \texttt{plot.dispRity} function graphically represents the summarised results using different representations: (1) ``continuous'' for displaying continuous disparity curves and (2) ``box'', ``lines'', or ``polygons'' to display them using boxplots, confidence interval lines or polygons respectively.
Additional arguments specific to \disp objects can also be used such as \texttt{observed} to display the observed disparity (i.e. non-bootstrapped) or \texttt{rarefaction} to only plot the disparity for a certain number of elements (i.e. the rarefaction level).
The function can also take any additional graphic arguments (\texttt{main}, \texttt{xlab}, \texttt{col}, etc...) from base \texttt{R}.

\subsubsection{Testing hypotheses}
The \texttt{test.dispRity} function allows users to test hypotheses on the disparity data.
Similarly to the \disp function described above, this function can take any test defined by the user or from other R packages.
The \texttt{comparison} arguments indicates in which order (if any) the tests should be applied to the subsets: (1) ``pairwise'' for pairwise comparisons; (2)``referential'' for comparing the first subset to all the others; (3) ``sequential'' for comparing subsets sequentially (e.g. first against second, second against third, etc.); (4) ``all'' for comparing all the subsets simultaneously (i.e. \texttt{disparity $\mathtt{\sim}$ subsets}) or (5) any list of pairs of subsets to compare.
